Boosting, Sparsity, and Conditional Gradient: A Tutorial

### 2014-03-10 arXiv/sparsity-result.tex
added a bit more exposition

Our sparsity result is algorithmic in nature, and relies on an algorithm called 
\emph{composite mirror descent} \cite{comid}. In composite mirror descent, we 
are given a sequence of functions $f_t(w) = g_t(w) + h_t(w)$, and the goal is to 
minimize
\[ \sum_{t=1}^T \alpha_t f_t(w), \]
for some known non-negative weights $\alpha_t$. We assume throughout that $f_t$, 
$g_t$, and $h_t$ are all convex. The composite mirror descent algorithm also 
depends on an auxiliary convex function $\psi(w)$:
\begin{algorithm}
\caption{The composite mirror descent algorithm.}
\label{alg:comid}
\begin{algorithmic}
\STATE {\bfseries Composite Mirror Descent}
\STATE Initialize $\theta_0 \gets 0$, $\psi_0 \gets \psi$
\FOR{$t = 1$ {\bfseries to} $T$}
  \STATE $\psi_t \gets \psi_{t-1} + \alpha_t h_t$
  \STATE $w_t \gets \argmin_{w \in \W} \psi_{t}(w) + \theta_{t-1}^{\top} w$
  \STATE $\theta_t \gets \theta_t + \alpha_t \nabla g_t(w_t)$
\ENDFOR
\STATE Output $w^* = \frac{\sum_{t=1}^T \alpha_t w_t}{\sum_{t=1}^T \alpha_t}$
\end{algorithmic}
\end{algorithm}

@@

% provide convergence theorem
We are now ready to state the convergence rate of Algorithm~\ref{alg:comid}.
\begin{theorem}
\label{thm:comid}
Let $\|\cdot\|$ be any norm on $\W$. Suppose that $\psi$ is $\rho_0$-strongly convex 
with respect to $\|\cdot\|$, and that $h_t$ is $\rho_t$-strongly convex with respect 
to $\|\cdot\|$. Suppose that $\|\nabla g_t\|_* \leq r$ for all $t$. Let 
$A_t = \sum_{s=1}^t \alpha_s$ and $R_t = \rho_0 + \sum_{s=1}^t \alpha_s \rho_s$. 
Then, defining $f(w) \eqdef \frac{1}{A_T} \sum_{t=1}^{T} \alpha_t f_t(w)$, we have
\[ f(w^*) \leq \inf_{w} f(w) + \frac{r^2}{2A_T} \sum_{t=1}^T \frac{\alpha_t^2}{R_t}. \]
\end{theorem}
The proof is given in Appendix~\ref{apx:comid}.

@@

\begin{corollary}
Suppose that $\psi$ is $\rho$-strongly convex with respect to $\|\cdot\|$, but make 
no assumptions on $h_t$ other than convexity. Then, for $\alpha_t = 1$, we have
\[ f(w^*) \leq \inf_w f(w) + \frac{r^2}{2\rho}. \]
\end{corollary}
\begin{proof}
We have $A_T = T$ and $R_t = \rho$. Then Theorem~\ref{thm:comid} yields
\begin{align*}
f(w^*) &\leq \inf_w f(w) + \frac{r^2}{2T} \sum_{t=1}^T \frac{1}{\rho} \\
 &= \inf_w f(w) + \frac{r^2}{2\rho}.
\end{align*}
\end{proof}

% give boosted version (cite Bach, Shai)
\subsection{Composite Mirror Descent and Sparsity}
\label{sec:boosted}
What relationship does Composite Mirror Descent have to sparsification? The answer lies
in a technique called \emph{boosting} which will allow us to cleverly pick the functions 
$f_t(w)$. Boosting is an ubiquitous idea spanning multiple literatures; % TODO provide citations
in our case, we suppose that we have fixed functions $h : \W \to \bR$ and 
$f : \V \times \W \to \bR$ such that:
\begin{enumerate}
\item $f(\cdot, w)$ is linear in $v$ for each $w \in \W$.
\item $f(v, \cdot) - h(\cdot)$ is convex in $w$ for each $v \in \V$.
\end{enumerate}
Let $v_t = \argmax_{v \in V} f(v, w_t)$. Then we will pick 
$h_t(w) = h(w)$ and $g_t(w) = f(v_t, w) - h(w)$. Note that, 
even though $g_t$ depends on $v_t$ and hence $w_t$, we can still 
run Algorithm~\ref{alg:comid}, since we do not need access to $g_t$ 
until \emph{after} selecting $w_t$.

This leads to the following primal-dual algorithm:
\begin{algorithm}
\caption{Primal-dual Composite Mirror Descent}
\label{alg:boosted-comid}
\begin{algorithmic}
\STATE Initialize $\theta_0 \gets 0$, $\psi_0 \gets \psi$
\FOR{$t = 1$ {\bfseries to} $T$}
  \STATE $\psi_t \gets \psi_{t-1} + \alpha_t h$
  \STATE $w_t \gets \argmin_{w \in \W} \psi_t(w) + \theta_{t-1}^{\top} w$
  \STATE $v_t \gets \argmax_{v \in \V} f(v, w_t)$
  \STATE $\theta_t \gets \theta_t + \alpha_t \left(\frac{\partial f(v_t, w_t)}{\partial w_t} - \frac{\partial h(w_t)}{\partial w_t}\right)$
\ENDFOR
\STATE Output $(v^*, w^*) = \left(\frac{\sum_{t=1}^T \alpha_t v_t}{\sum_{t=1}^T \alpha_t}, \frac{\sum_{t=1}^T \alpha_t w_t}{\sum_{t=1}^T \alpha_t}\right)$
\end{algorithmic}
\end{algorithm}

% give convergence theorem for boosted version
Algorithm~\ref{alg:boosted-comid} enjoys the following primal-dual convergence rate:
\begin{theorem}
TODO
\end{theorem}

>>>>

Our sparsity result is algorithmic in nature, and relies on an algorithm called 
\emph{composite mirror descent} \cite{comid}. In composite mirror descent, we 
are given a sequence of functions $f_t(w) = g_t(w) + h_t(w)$, and the goal is to 
minimize
\[ \sum_{t=1}^T \alpha_t f_t(w), \]
for some known non-negative weights $\alpha_t$. We assume throughout that $f_t$, 
$g_t$, and $h_t$ are all convex. The composite mirror descent algorithm also 
depends on an auxiliary convex function $\psi(w)$:
\begin{algorithm}
\caption{Composite Mirror Descent}
\label{alg:comid}
\begin{algorithmic}
\STATE Initialize $\theta_0 \gets 0$, $\psi_0 \gets \psi$
\FOR{$t = 1$ {\bfseries to} $T$}
  \STATE $\psi_t \gets \psi_{t-1} + \alpha_t h_t$
  \STATE $w_t \gets \argmin_{w \in \W} \psi_{t}(w) + \theta_{t-1}^{\top} w$
  \STATE $\theta_t \gets \theta_t + \alpha_t \nabla g_t(w_t)$
\ENDFOR
\STATE Output $w^* = \frac{\sum_{t=1}^T \alpha_t w_t}{\sum_{t=1}^T \alpha_t}$
\end{algorithmic}
\end{algorithm}

@@

% provide convergence theorem
We are now ready to state the convergence rate of Algorithm~\ref{alg:comid}.
\begin{theorem}
\label{thm:comid}
Let $\|\cdot\|$ be any norm on $\W$. Suppose that $\psi$ is non-negative and $\rho_0$-strongly convex 
with respect to $\|\cdot\|$, and that $h_t$ is $\rho_t$-strongly convex with respect 
to $\|\cdot\|$. Suppose that $\|\nabla g_t\|_* \leq r$ for all $t$. Let 
$A_t = \sum_{s=1}^t \alpha_s$ and $R_t = \rho_0 + \sum_{s=1}^t \alpha_s \rho_s$. 
Then, defining $f(w) \eqdef \frac{1}{A_T} \sum_{t=1}^{T} \alpha_t f_t(w)$, we have
\[ f(w^*) \leq \inf_{w} \left\{ f(w) + \frac{1}{A_T} \psi(w)\right\} + \frac{r^2}{2A_T} \sum_{t=1}^T \frac{\alpha_t^2}{R_t}. \]
\end{theorem}
The proof is given in Appendix~\ref{apx:comid}.

@@

\begin{corollary}
Suppose that $\psi$ is $\rho_0$-strongly convex with respect to $\|\cdot\|$, but make 
no assumptions on $h_t$ other than convexity. Then, for $\alpha_t = 1/\sqrt{T}$, we have
\[ f(w^*) \leq \inf_w \left\{ f(w) + \frac{1}{\sqrt{T}} \psi(w)\right\} + \frac{r^2}{2\rho_0\sqrt{T}}. \]
\end{corollary}
\begin{proof}
We have $A_T = \sqrt{T}$ and $R_t = \rho_0$. Then Theorem~\ref{thm:comid} yields
\begin{align*}
f(w^*) &\leq \inf_w \left\{ f(w) + \frac{1}{\sqrt{T}} \psi(w)\right\} + \frac{r^2}{2\sqrt{T}} \sum_{t=1}^T \frac{1}{\rho_0 T} \\
 &= \inf_w \left\{ f(w) + \frac{1}{\sqrt{T}} \psi(w)\right\} + \frac{r^2}{2\rho_0\sqrt{T}}.
\end{align*}
\end{proof}
So, if the $h_t$ are strongly convex, we can obtain an $O\left(\frac{1}{T}\right)$ 
convergence rate. If the $h_t$ are only (weakly) convex, we can still obtain an 
$O\left(\frac{1}{\sqrt{T}}\right)$ convergence rate, by ``deforming'' $f$ by a 
strongly convex function $\psi$.

% give boosted version (cite Bach, Shai)
\subsection{Composite Mirror Descent and Sparsity}
\label{sec:boosted}
What relationship does Composite Mirror Descent have to sparsification? The answer lies
in a technique called \emph{boosting} which will allow us to cleverly pick the functions 
$f_t(w)$. Boosting is an ubiquitous idea spanning multiple literatures; % TODO provide citations
in our case, we suppose that we have fixed functions $h : \W \to \bR$ and 
$f : \V \times \W \to \bR$ such that:
\begin{enumerate}
\item $f(\cdot, w)$ is linear in $v$ for each $w \in \W$.
\item $f(v, \cdot) - h(\cdot)$ is convex in $w$ for each $v \in \V$.
\end{enumerate}
We will also assume that there is a compact set $S \subset \V$ with respect 
to which we are trying to obtain sparsity. Let 
$v_t = \argmax_{v \in S} f(v, w_t)$. Then we will pick 
$h_t(w) = h(w)$ and $g_t(w) = f(v_t, w) - h(w)$. Note that, 
even though $g_t$ depends on $v_t$ and hence $w_t$, we can still 
run Algorithm~\ref{alg:comid}, since we do not need access to $g_t$ 
until \emph{after} selecting $w_t$.

This leads to the following primal-dual algorithm:
\begin{algorithm}
\caption{Primal-dual Composite Mirror Descent}
\label{alg:boosted-comid}
\begin{algorithmic}
\STATE Initialize $\theta_0 \gets 0$, $\psi_0 \gets \psi$
\FOR{$t = 1$ {\bfseries to} $T$}
  \STATE $\psi_t \gets \psi_{t-1} + \alpha_t h$
  \STATE $w_t \gets \argmin_{w \in \W} \psi_t(w) + \theta_{t-1}^{\top} w$
  \STATE $v_t \gets \argmax_{v \in S} f(v, w_t)$
  \STATE $\theta_t \gets \theta_t + \alpha_t \left(\frac{\partial f(v_t, w_t)}{\partial w_t} - \frac{\partial h(w_t)}{\partial w_t}\right)$
\ENDFOR
\STATE Output $(v^*, w^*) = \left(\frac{\sum_{t=1}^T \alpha_t v_t}{\sum_{t=1}^T \alpha_t}, \frac{\sum_{t=1}^T \alpha_t w_t}{\sum_{t=1}^T \alpha_t}\right)$
\end{algorithmic}
\end{algorithm}

% give convergence theorem for boosted version
Algorithm~\ref{alg:boosted-comid} enjoys the following primal-dual convergence rate:
\begin{theorem}
Define the \emph{duality gap} $\Delta_T$ as
\[ \Delta_T \eqdef \sup_{v \in \Conv(S)} f(v, w_T) - \inf_{w \in \W} f(v_T, w). \]
Suppose that $\psi$ is $\rho_0$-strongly convex with 
respect to a norm $\|\cdot\|$, and that $h$ is $\rho$-strongly 
convex with respect to the same norm. Suppose that 
$\|\nabla g_t\|_* \leq r$ for all $t$. Then, defining 
$A_t \eqdef \sum_{s=1}^T \alpha_s$ as before, we have
\[ \Delta_T \leq \frac{1}{A_T} \psi(w_T) + \frac{r^2}{2A_T} \sum_{t=1}^T \frac{\alpha_t^2}{\rho_0 + \rho A_t}. \]
In addition, $v_T$ is $(S, T)$-sparse.
\end{theorem}
%\begin{remark}
%If $f$ is a continuously differentiable function, the compactness of $S$ 
%guarantees that $\sup_{v \in S} \|\nabla g(v, w)\|_*$ is finite for each 
%$w$. 
%\end{remark}
\begin{corollary}
Suppose that $\psi \equiv 0$ and $\rho > 0$, and set $\alpha_t = t$. Then 
\[ \Delta_T \leq \frac{2r^2}{\rho (T+1)}. \]
\end{corollary}

\begin{corollary}
Suppose that $\rho_0 > 0$ and set $\alpha_t = 1/\sqrt{T}$. Then
\[ \Delta_T \leq \frac{1}{\sqrt{T}} \left[ \psi(w_T) + \frac{r^2}{2\rho_0} \right]. \]
\end{corollary}
